---
title: "Checking for Moderators in Linear Mixed Effects Models with Longitudinal Data"
output: html_document
date: "2025-01-24"
---

# Load packages
```{r}
library(tidyverse)
library(broom.mixed)
library(lmerTest)
library(emmeans)
```


# View example dataset
```{r}
head(sleepstudy)
```

# Assign treatment groups and moderator
```{r}
sleepstudy <- sleepstudy %>% 
  filter(Days <= 1) %>% 
  mutate(treatment = case_when(
    as.numeric(Subject) < 10 ~ 0, 
    as.numeric(Subject) >= 10 ~ 1,
    TRUE ~ NA
  ),
  sex = as.factor(case_when(
    as.numeric(Subject) %% 2 == 0 ~ "Male",
    TRUE ~ "Female"
  )),
  Days = as.factor(Days))
```


# Fit overall model - interaction between time * treatment
```{r}
m1 <- lmer(Reaction ~ treatment * Days + (1|Subject), data = sleepstudy)
```


# Get time * treatment effect estimate, 95% CI, p-value
```{r}
m1_summary <- tidy(m1, conf.int = TRUE) 

m1_summary
```

# Check for moderation by sex (screen for significant 3-way interaction)
```{r}
m2 <- lmer(Reaction ~ treatment * Days * sex + (1|Subject), data = sleepstudy)
```


# Get time * treatment * sex effect estimate, 95% CI, p-value
```{r}
m2_summary <- tidy(m2, conf.int = TRUE) 

m2_summary
```

# If p-value significant, then move on to post-hoc estimation of effects by group
```{r}
emmeans(m2, specs = c("treatment", "Days", "sex")) %>% 
  contrast(interaction = "consec", by = "sex") %>% 
  tidy()
```

