---
title: "Checking for Moderators in Linear Mixed Effects Models"
output: html_document
date: "2025-03-28"
---

# Load packages
```{r}
library(tidyverse)
library(broom.mixed)
library(lmerTest)
library(emmeans)
```

# View example dataset
```{r}
glimpse(carrots)
```

# Fit overall model with exposure = sweetness and outcome = preference including random effect for consumer
```{r}
m1 <- lmer(Preference ~ Sweetness + (1|Consumer), data = carrots)
```


# Get effect (beta) of sweetness on preference, 95% CI, p-value
```{r}
m1_summary <- tidy(m1, conf.int = TRUE) 

m1_summary
```

# Check for moderation by Gender (screen for significant 2-way interaction)
```{r}
m2 <- lmer(Preference ~ Sweetness * Gender + (1|Consumer), data = carrots)
```


# Check for statistical interaction (Gender * Sweetness), 95% CI, p-value
```{r}
m2_summary <- tidy(m2, conf.int = TRUE) 

m2_summary
```

# Post-hoc estimation of effects of sweetness on preference by Gender
```{r}
# Use this to get betas for Sweetness across Genders
# First argument is a fitted model object
# specs= the name of a variable across whose levels betas will be estimeated
# var= the variable whose betas we want to estimate
emtrends(m2, specs = "Gender", var = "Sweetness") %>% 
  tidy(conf.int = TRUE)
```
# For more info, see this workshop: https://stats.oarc.ucla.edu/wp-content/uploads/2023/03/interactions_r.html#(1)

