---
title: Performing DID using linear mixed effects models applied to multiply imputed
  data
output: html_document
date: "2025-05-22"
---

```{r}
library(tidyverse)
library(emmeans) 
library(mice)
library(ggmice)
library(broom.mixed)
library(lme4)
library(collapse)
```

# Simulate dataset with characteristics we need
```{r}
sim_data <- tibble(
  subject = c(1:100),
  group = rep(c(1,0), each = 50),
  score_t1 = runif(100, min = 0, max = 100),
  sex = sample(c("M", "F"), 100, replace = TRUE),
  age = sample(18:65, 100, replace = TRUE),
  race = sample(c("white", "black", "other"), 100, replace = TRUE)
)

sim_data <- sim_data %>% 
  # Generate score_t2 for each group
  mutate(score_t2 = case_when(
    group == 1 ~ score_t1 + sample(0:10, n(), replace = TRUE),
    group == 0 ~ score_t1 - sample(0:5, n(), replace = TRUE)
  ),
  # Introduce some random missing values for score_t2
  score_t2 = case_when(
    runif(n()) < 0.15 ~ NA_real_, TRUE ~ score_t2
  ))
```

# Reformat the data for multiple imputation, character variables must be factors
```{r}
sim_data_to_impute <- sim_data %>% 
  mutate(sex = as.factor(sex),
         race = as.factor(race))
```

# Convert data to long
```{r}
sim_data_to_impute_long <- sim_data_to_impute %>% 
  pivot_longer(cols = c("score_t1", "score_t2"), names_to = "time", values_to = "score", names_prefix = "score_") %>% 
  mutate(time = as.factor(time))
```

# Define the longitudinal multiple imputation model
```{r}
imp0 <- mice(sim_data_to_impute_long, maxit=0)
meth <- imp0$method
predM <- imp0$predictorMatrix

# Define score as requiring a longitudinal/time-dependent predictive model
meth["score"] <- "2l.norm"

# In predictor matrix, set subject variable to -2 to treat as grouping variable, and time as 2 for a random effect/repeated measure
predM[, "subject"] <- -2
predM[, "time"] <- 2
```


# Check the predictor matrix
```{r}
predM
```
# Confirm that the predictive model looks appropriate based on the data type
```{r}
meth
```
# Now let's impute!
```{r}
imp <- mice(sim_data_to_impute_long, m = 5, seed = 111111, print = TRUE,
            predictorMatrix = predM, method = meth)
```

# Check for logged events
```{r}
imp$loggedEvents
```

# Check for convergence
```{r}
plot(imp)
```

# Summarize the imputation data for visualizations, etc.
```{r}
imp_data_long <- complete(imp, "long", include = FALSE)

num_summary <- imp_data_long %>% 
  dplyr::select(subject, time, where(is.numeric)) %>% 
  group_by(subject, time) %>% 
  summarise_all(mean)

fct_summary <- imp_data_long %>% 
  dplyr::select(subject, time, where(is.factor)) %>% 
  group_by(subject, time) %>% 
  summarise_all(fmode)

imp_summary <- num_summary %>% 
  full_join(fct_summary, by = c("subject", "time"))
```

# Check t1 and t2 distributions of score, averaged across all imputations
```{r}
imp_summary %>% 
  ggplot(mapping = aes(x = score)) +
  geom_histogram(bins = 15) +
  facet_grid(rows = vars(group), cols = vars(time))
```

# Unadjusted means and DID estimate
```{r}
fit <- with(data = imp,
            exp = lmer(score ~ group * time + (1|subject)))

m1 <- pool(fit)

as.data.frame(summary(m1, conf.int = TRUE))
```

# Within-group change (post-pre)
```{r}
emmeans(fit, specs = c("group", "time"), adjust = "none") %>% 
  pairs(reverse = TRUE, adjust = "none") %>% 
  confint() %>% 
  filter(contrast %in% c("group0 t2 - group0 t1", "group1 t2 - group1 t1")) %>% 
  as_tibble()
```

# Group means
```{r}
emmeans(fit, specs = c("group", "time"))
```
# Adjusted DID estimate
```{r}
fit_adj <- with(data = imp,
                exp = lmer(score ~ group * time +
                             age + sex + race + (1|subject)))

m1_adj = pool(fit_adj)

as.data.frame(summary(m1_adj, conf.int = TRUE))
```

# Plot results
```{r}
# Estimated means and standard error for each group and time
m1_means <- tidy(emmeans(fit, specs = c("group", "time"))) %>% 
  dplyr::select(group, time, estimate, std.error)

m1_plot_data <- m1_means %>% 
  mutate(group = factor(case_when(
    group == 1 ~ "Intervention",
    group == 0 ~ "Control"
  ), levels = c("Intervention", "Control")),
    lower = estimate + (-1.96 * std.error),
    upper = estimate + (1.96 * std.error))

m1_plot <- m1_plot_data %>% 
  ggplot(mapping = aes(x = time, y = estimate, color = group)) +
  geom_line(aes(group = group), linewidth = 1.5, linetype = "solid") +
  geom_pointrange(aes(ymin = lower, ymax = upper), fatten = 10, linewidth = 1.5) +
  theme_bw() +
  labs(y = "Mean Score",
       x = "Time",
       color = "Treatment Group") + 
  ylim(0, 65)

m1_plot
```

